A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-21 06-appendix.tex
Correção Rodrigo e remoção do "Problema"

  \begin{case}
    $p=\ceu{\Ifelse{\Mem(\Id)}{p'}{p''}}$.  There are two subcases.
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$ is false.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p';\,p''}$.  There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$ and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

@@

\begin{theorem}\label{thm.term-nst-*}
  For any~$\delta$ there is a~$\delta'_\Hnst$ such
  that~$\delta\nst[*]\delta'_\Hnst$.
\end{theorem}
\begin{proof}
  By induction on the structure of programs.
  %%
  Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
  nested-irreducible, as by definition~$\delta\nst[0]\delta_\Hnst$.  Suppose
  that is not the case.  Then, depending on the structure of~$p$, there
  are~11 possibilities.  In each one of them, we show that
  such~$\delta'_\Hnst$ indeed exists.
  \begin{case}
    $p=\ceu{\Mem(\Id)}$.
    %%
    Then, by~\R{mem},
    \[
      \<\ceu{\Mem(\Id)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
    \]
  \end{case}

@@

% LocalWords:  subcase subcases

>>>>

  \begin{case}
    $p=\ceu{\IfElse{\Mem(\Id)}{p'}{p''}}$.
    %%
    There are two subcases.
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-true}.
      Hence~$p_1=p_2=p'$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$ is false.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{if-false}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p';\,p''}$.
    %%
    There are three subcases.
    \begin{subcase}
      $p'=\ceu{\Nop}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-nop}.
      Hence~$p_1=p_2=p''$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'=\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-brk}.
      Hence~$p_1=p_2=\ceu{\Break}$ and~$e_1=e_2=\nil$.
    \end{subcase}
    \begin{subcase}
      $p'\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi_1$ and~$\pi_2$ are instances of~\R{seq-adv}.
      Thus there are derivations~$\pi_1'$ and~$\pi_2'$ such that
      \begin{align*}
        \pi_1'&\Vdash\<p',n,\nil>\nst\<p_1',n,e_1'>\\
        \pi_2'&\Vdash\<p',n,\nil>\nst\<p_2',n,e_2'>
      \end{align*}
      for some~$p_1',p_2'\in\P$ and~$e_1',e_2'\in\E$.  By the induction
      hypothesis, $p_1'=p_2'$ and~$e_1'=e_2'$.
      Hence~$p_1=\ceu{p_1';p''}=\ceu{p_2';p''}=p_2$ and~$e_1=e_1'=e_2'=e_2$.
    \end{subcase}
  \end{case}

@@

\begin{theorem}\label{thm.term-nst-*}
  For any~$\delta$ there is a~$\delta'_\Hnst$ such
  that~$\delta\nst[*]\delta'_\Hnst$.
\end{theorem}
\begin{proof}
  By induction on the structure of programs.
  %%
  Let~$\delta=\<p,n,\nil>$.  The theorem is trivially true if~$\delta$ is
  nested-irreducible, as by definition~$\delta\nst[0]\delta$.  Suppose that
  is not the case.  Then, depending on the structure of~$p$, there are~11
  possibilities.  In each one of them, we show that such~$\delta'_\Hnst$
  indeed exists.
  \begin{case}
    $p=\ceu{\Mem(\Id)}$.
    %%
    Then, by~\R{mem},
    \[
      \<\ceu{\Mem(\Id)},n,\nil>\nst[1]\<\ceu{\Nop},n,\nil>_\Hnst\,.
    \]
  \end{case}

@@

\begin{definition}\label{def.pot}
  The potency of a program~$p$ in reaction to event~$e$,
  denoted~$\pot(p,e)$, is the maximum number of~$\ceu{\EmitInt}$ expressions
  that can be reached during a reaction of~$p$ to~$e$.  Formally,
  $\pot(p,e)$ is defined as follows:
  \[
    \pot(p,e)=\pot'(\bcast(p,e))\,,
  \]
  where
  \begin{enumerate}[(a)]
  \item$\pot'(p)=1$, if~$p=\ceu{\EmitInt(e)}$;
    %%
  \item$\pot'(p)=\max\{\pot'(p_1),\pot'(p_2)\}$,\\
    if~$p=\ceu{\IfElse{\Mem(\Id)}{p_1}{p_2}}$;
    %%
  \item$\pot'(p)=\pot'(p_1)$, if~$p=\ceu{\Loop{p_1}}$
    or~$p=\ceu{\Every{e}\,{p_1}}$;
    %%
  \item$\pot'(p)=\pot'(p_1)+\pot'(p_2)$, if~$p=p_1;\,p_2$
    or~$p=\ceu{p_1\AtLoop{p_2}}$
    with~$p_1\ne\ceu{\Break},\ceu{\AwaitExt(e)}$,
    or~$p=\ceu{{p_1}\And{p_2}}$ or~$p=\ceu{{p_1}\Or{p_2}}$
    or~$p=\ceu{{p_1}\AtAnd{p_2}}$ or~$p=\ceu{{p_1}\AtOr{p_2}}$;
    %%
  \item$\pot'(p)=0$, otherwise.
  \end{enumerate}
  Note that~$\pot'$ only considers the~$\ceu{\EmitInt}$ expressions that
  occur before the first~$\ceu{\AwaitExt}$ in a given execution path.  This
  is because such~$\ceu{\AwaitExt}$ act as barriers that cannot be consumed
  in the reaction to the original event~$e$.
\end{definition}

\begin{definition}\label{def.rank}
  The \emph{rank} of a description~$\delta=\<p,n,e>$,
  denoted~$\rank(\delta)$, is a pair of nonnegative integers~$\<i,j>$ such
  that~$i=\pot(p,e)$, and $j=n$~if~$e=\nil$ or~$j=n+1$ otherwise.
\end{definition}

The next two lemmas establish that a single application of an~$\out$
or~$\nst$ transition either preserves or decreases the rank of the input
description.  All rank comparisons assume lexicographical order.

\begin{lemma}\strut
  \begin{enumerate}[(a)]
  \item If~$\delta\outpush\delta'$ then~$\rank(\delta)=\rank(\delta')$.
  \item If~$\delta\outpop\delta'$ then~$\rank(\delta)>\rank(\delta')$.
  \end{enumerate}
\end{lemma}
\begin{proof}\strut
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$,
  $\rank(\delta)=\<i,j>$, and~$\rank(\delta')=\<i',j'>$.
  \begin{enumerate}[(a)]
  \item Suppose
    \[
      \<p,n,e>\outpush\<p',n',e'>\,.
    \]
    By~\R{push}, $e\ne\nil$, $p'=\bcast(p,e)$, $n'=n+1$, and~$e'=\nil$.  By
    Definition~\ref{def.rank}, $j=n+1$, as~$e\ne\nil$, and~$j'=n+1$,
    as~$e'=\nil$ and~$n'=n+1$; hence~$j=j'$.

    It remains to be shown that~$i=i'$:
    \begin{align*}
      i&=\pot(p,e)
         \tag*{by Definition~\ref{def.rank}}\\
       &=\pot'(\bcast(p,e))
         \tag*{by Definition~\ref{def.pot}}\\
       &=\pot'(p')
         \tag*{since~$p'=\bcast(p,e)$}\\
       &=\pot'(\bcast(p',\nil))
         \tag*{by definition of~$\bcast$}\\
       &=\pot'(\bcast(p',e'))
         \tag*{since~$e'=\nil$}\\
       &=\pot(p',e')
         \tag*{by Definition~\ref{def.pot}}\\
       &=i'
         \tag*{by Definition~\ref{def.rank}}
    \end{align*}

  \item Suppose
    \[
      \<p,n,e>\outpop\<p',n',e'>\,.
    \]
    By~\R{pop}, $p=p'$, $n>0$, $n'=n-1$, and~$e=e'=\nil$.
    %%
    By Definition~\ref{def.pot}, $\pot(\bcast(p,e))=\pot(\bcast(p',e')$;
    hence~$i=i'$.  And by Definition~\ref{def.rank}, $j=n$, as~$e=\nil$,
    and~$j'=n-1$, as~$e'=\nil$ and~$n'=n-1$; hence~$j>j'$.
    Therefore,~${\<i,j>}>{\<i',j'>}$.\qedhere
  \end{enumerate}
\end{proof}

\begin{lemma}\label{lem.rank-nst}
If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.
  Let~$\delta=\<p,n,e>$, $\delta'=\<p',n',e'>$, $\rank(\delta)=\<i,j>$,
  and~$\rank(\delta')=\<i',j'>$.  Suppose there is a derivation~$\pi$
  such that
  \[
    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  \]
  By definition of~$\nst$, we have that~$e=\nil$ and $n=n'$.  Depending on
  the structure of program~$p$, there are~11 possibilities.  In each one of
  them, we show that~$\rank(\delta)\ge\rank(\delta')$.

  \begin{case}
    $p=\ceu{\Mem(id)}$.
    %%
    Then~$\pi$ is an instance of~\R{mem}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
  \end{case}

  \begin{case}
    $p=\ceu{\EmitInt(e_1)}$.
    %%
    Then~$\pi$ is an instance of~\R{emit-int}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=e_1$.
    Thus~${\rank(\delta)={\<1,n>}}>{\<0,n+1>=\rank(\delta')}$.
  \end{case}

  \begin{case}
    $p=\ceu{\CanRun(n)}$.
    %%
    Then~$\pi$ is an instance of~\R{can-run}.  Hence~$p'=\ceu{\Nop}$
    and~$e'=\nil$.  Thus~$\rank(\delta)=\rank(\delta')=\<0,n>$.
  \end{case}

  \begin{case}
    $p=\ceu{\IfElse{p}{p_1}{p_2}}$.
    %%
    There are two subcases.
    \begin{subcase}
      \label{lem.rank-nst.if-true}
      $\eval(\ceu{\Mem(\Id)})$~is true.
      %%
      Then~$\pi$ is an instance of~\R{if-true}.  Hence~$p'=\ceu{p_1}$
      and~$e'=\nil$.  Thus~$\rank(\delta)=\<i,n>$
      and~$\rank(\delta')=\<j,n>$, where~$i=\max\{pot'(p_1),pot'(p_2)\}$
      and~$j=\pot'(p_1)$.  Therefore~$i>j$.
    \end{subcase}
    \begin{subcase}
      $\eval(\ceu{\Mem(\Id)})$~is false.
      %%
      Similar to Case~\ref{lem.rank-nst.if-true}.
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{p_1;\,p_2}$.
    %%
    There are three subcases.
    \begin{subcase}
      $p_1=\ceu{\Nop}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-nop}.  TODO.
    \end{subcase}
    \begin{subcase}
      $p_1=\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-brk}.  TODO.
    \end{subcase}
    \begin{subcase}
      \label{lem.rank-nst.seq-adv}
      $p_1\ne\ceu{\Nop},\ceu{\Break}$.
      %%
      Then~$\pi$ is an instance of~\R{seq-adv}.
      Thus there is a derivation~$\pi'$ such that
      \[
        \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
      \]
      for some~$p_1'\in\P$ and~$e_1'\in\E$.  Thus~$p'=p_1';p_2$
      and~$e'=e_1'$.  By the induction hypothesis,
      \begin{equation}
        \label{lem.rank-nst.seq-adv.eq1}
        \rank(\<p_1,n,\nil>)\ge\rank(\<p_1',n,e_1'>)\,.
      \end{equation}
      There are two subcases.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then
        \begin{align*}
          \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\,,\enspace\text{and}\\
          \rank(\delta')&=\<\pot'(p_1')+\pot'(p_2),n>\,.
        \end{align*}
        By~\eqref{lem.rank-nst.seq-adv.eq1} and from the fact
        that~$e_1'=\nil$, $\pot'(p_1)\ge\pot'(p_1')$.  Therefore,
        $\rank(\delta)\ge\rank(\delta')$.
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ is an instance of~\R{emit-int}, as no other rule affects
        the third component of the output description.
        \[
          \rank(\delta)=\<i,n>
          >\<i-1,n+1>
          =\rank(\delta')\,.
        \]
      \end{subsubcase}
    \end{subcase}
  \end{case}

  \begin{case}
    $p=\ceu{\Loop{p_1}}$.
  \end{case}

  TODO\dots
\end{proof}

\begin{theorem}[Reaction termination]
  For any~$\delta$ there is a~$\delta'_\Hnst$ such
  that~$\delta\trans[*]\delta'$ and~$\rank(\delta')=\<0,0>$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$rank(\delta)$.
\end{proof}
%   \begin{basis}
%     If~$\<i,j>=\<0,0>$ then, by~\ref{?}, $\delta$ cannot be advanced
%     by~$\out$ transitions.  There are two possibilities\dots\ (it can or
%     cannot be advanced by nst transitions).
%   \end{basis}
%   \begin{induction}
%     Let~$\<i,j>\ne\<0,0>$ and suppose that the theorem holds for
%     all~$\delta''$ such that~$\rank(\delta'')<\rank(\delta)$.
%     \begin{case}
%       $\delta$~is nested-irreducible.
%       %%
%       \begin{subcase}
%         $\delta\outpush\delta_1$, for some~$\delta_1$.
%       \end{subcase}
%       \begin{subcase}
%         $\delta\outpop\delta_1$, for some~$\delta_1$.
%         %%
%         Then, by definition of~\R{pop}, $\rank(\delta_1)=\<i,j-1>$, and by
%         the induction hypothesis\dots
%       \end{subcase}
%     \end{case}
%     \begin{case}
%       $\delta$~is not nested-irreducible.
%       %%
%     \end{case}
%   \end{induction}
